Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2012-09-05 Dail_Mad_ext.tex
end of day

Theoretical ecology requires models of population dynamics for testing
hypotheses regarding spatial and temporal variation in
abundance---hypotheses relating to the importance and existence of
phenomenon such as density-dependent population regulation, population
cycling, and spatial synchrony
\citep{may:1975,royama:1977,turchin:1990,dennis_taper:1994,bjornstad_etal:1999}.
In applied contexts, population models are used for
estimating extinction probabilities
\citep{schoener_spiller:1992,nadeem_lele:2011} and for predicting the
effects of future environmental conditions or conservation actions on
population size \citep{jamieson_brooks:2004,hatfield_etal:2012}.
In order to address these questions, several complicating factors must
be confronted when fitting population models to data.
First, deterministic models of population
dynamics are virtually always inadequate due to process variation,
the inherent stochasticity in demographic parameters and environmental
conditions \citep{bjornstad_grenfell:2001}. %clark_bjornstad:2004}.
Second, abundance---the natural state variable in studies
of population dynamics---can rarely be observed perfectly in field
studies because of observation error, such as imperfect
detection \citep{link_nichols:1994,kery_etal:2009}.
%Failure to account for process
%variation and observation error can bias the estimators of abundance
%and related parameters.

@@

One of the simplest state-space models can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a deterministic population growth model
(e.g. exponential, logistic, etc \dots) and a stochastic model allowing
for random process variation, which we denote $\grave{o}_t$ . As an
example, under exponential population growth, the model could be
written as:
\begin{gather}
N_t = N_{t-1}\exp(r)+\grave{o}_t \nonumber \\
X_t = N_t + \zeta_t
\label{eq:ss1}
\end{gather}
where $r$ is the intrinsic rate of increase. Typically, both error terms
are assumed to be i.i.d random effects distrubted as
$\grave{o}_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. The Gaussian assumption allows for parameter
estimation using the Kalman filter \citep{dennis_etal:2006}, which is
much more computationally efficient than estimation methods when
random effects are not Gaussian distributed
\citep{devalpine_hastings:2002}.

Exponential growth without density-dependence is rarely observed in
natural populations, and thus $\exp(r)$ is often substituted for other
population growth models such as the Ricker or Gompertz. Nonetheless,
the basic structure shown in Eq.~\ref{eq:ss1} is common to most
traditional state-space models. However, even though this is
the most widespread approach for modeling population dynamics using
time-series data, several problems are immediately evident. First,
both $N_t$ and $X_t$ are typically non-negative integers, and hence
the the Gaussian assumption is not justified.
%is clearly a problem with this formulation because the normally distributed
%random effects violate this constraint.
One approach to circumventing this problem in hopes of adhering to the
Gaussian assumption is to tranform abundance to
density by replacing $N_t$ with $Y_t = \log(N_t/A)$ where $A$ is the
area surveyed. However, this transformation is problematic since
abundance may be zero and thus $Y_t = \log(0) = -\infty$. Generally,
zeros are replaced with some small number, although the
effect of this is rarely discussed.

@@

A third problem with these models is that they do not admit
spatial variation. This is not a fair criticism because this
lies beyond the scope of traditional state-space models, but
it does restrict the utility of their since it is typically
impossible assess the impacts of factors such as habitat
fragmentation or climate change on a population without
considering spatial variation. Indeed many populations are
regulated by spatial processes such as source-sink dynamics.

\citet{dail_madsen:2011} developed a model (henceforth the DM model)
that resolves each of these problems with traditional state-space
models. The DM model allows for inference about spatial and temporal
dynamics in abundance while accounting for observation error using
only spatially and temporally replicated count data.  Their model is
an extension of the closed-population N-mixture model
\citep{royle:2004biom}, which was designed specifically to address the
problem of
modeling spatial variation in abundance when detection probability is
less than unity. The DM extension relaxes the assumption of population
closure and includes explicit parameters describing population change
as a first-order Markovian process. [more details]

[applications / potential etc... ] The DM model is relevant to a huge
number of problems confronting basic and applied ecological
research. Answering questions about climate change etc... require
spatially and temporally extensive datasets such as result from the
North American Breeding Bird Survey (BBS)
\citep{robbins_etal:1986}. However, most of the datasets of the
appropriate scale are
collected by volunteers and using protocols that are not amenable to
traditional approaches of modeling abundance and detection probability
based on standard capture-recapture methods. Instead they have, at
best, simple count data for which, when it can be interpreted as
counts of unique individuals, the DM model can be applied.. Yadi ya

>>>>

Theoretical ecology requires models of population dynamics for testing
hypotheses regarding spatial and temporal variation in
abundance---hypotheses relating to the importance and existence of
phenomenon such as density-dependent population regulation, population
cycling, and spatial synchrony
\citep{may:1975,royama:1977,turchin:1990,dennis_taper:1994,bjornstad_etal:1999}.
In applied contexts, population models are used for
estimating extinction probabilities
\citep{schoener_spiller:1992,nadeem_lele:2011} and for predicting the
effects of future environmental conditions or conservation actions on
population size \citep{jamieson_brooks:2004,hatfield_etal:2012}.
In order to address these questions, two complicating factors must
be confronted when fitting population models to data.
First, deterministic models of population
dynamics are virtually always inadequate due to process variation,
the inherent stochasticity in demographic parameters and environmental
conditions \citep{bjornstad_grenfell:2001}. %clark_bjornstad:2004}.
Second, abundance---the natural state variable in studies
of population dynamics---can rarely be observed perfectly in field
studies because of observation error, such as imperfect
detection \citep{link_nichols:1994,kery_etal:2009}.
%Failure to account for process
%variation and observation error can bias the estimators of abundance
%and related parameters.

@@

One of the simplest state-space models can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a deterministic population growth model
(e.g. exponential, logistic, etc \dots) and a stochastic model allowing
for random process variation, which we denote $\grave{o}_t$. As an
example, under exponential population growth, the model could be
written as:
%\begin{gather}
%N_t = N_{t-1}\exp(r)+\grave{o}_t \nonumber \\
%X_t = N_t + \zeta_t
%\label{eq:ss1}
%\end{gather}
\begin{subequations}
  \label{eq:ss1}
  \begin{gather}
    \mu_t = N_{t-1}\exp(r) \label{eq:ss1a} \\ \label{eq:ss1b}
    N_t \sim \text{N}(\mu_t, \grave{o}_t) \\ \label{eq:ss1c}
    X_t \sim \text{N}(N_t, \zeta_t)
  \end{gather}
\end{subequations}
where $r$ is the intrinsic rate of increase. Although the exponential
growth model shown here is often replaced with density dependent
models, most state-space models share the common feature that the
error terms are assumed to be i.i.d random effects distrubted as
$\grave{o}_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. The Gaussian assumption allows for parameter
estimation using the Kalman filter \citep{dennis_etal:2006}, which is
much more computationally efficient than estimation methods when
random effects are not Gaussian distributed
\citep{devalpine_hastings:2002}.

Even though state-space models such as that shown in Eq.~\ref{eq:ss1}
are the most widespread approach for modeling population dynamics
using time-series data, several problems are evident. Namely, (1)
time-series data and the underlying abundance parameters of interest
are typically integer-valued, raising concerns about the use of
Gaussian distribution shown in Eq~\ref{eq:ss1b}, (2) the model for
observatoin error (Eq.~\ref{eq:ss1c}) has little biological basis, (3)
spatial variation in abundance is not allowed, and (4) some of the
paramters of the model are not estimable. We briefly discuss each of
these points before describing a general approach to resolving these
shortcomings.

The fact that both abundance ($N_t$) and the observed data ($X_t$) are
typically non-negative integers suggests that the
Gaussian assumption in Eq.~\ref{eq:ss1b} is not justified.
%is clearly a problem with this formulation because the normally distributed
%random effects violate this constraint.
Even so, efforts are often made to adhere to the
Gaussian assumption by tranforming abundance to
density by replacing $N_t$ with $Y_t = \log(N_t/A)$ where $A$ is the
area surveyed. However, this transformation is problematic since
abundance may be zero and thus $Y_t = \log(0) = -\infty$. Generally,
zeros are replaced with some small number, although the
effect of this is rarely discussed.

Another problem with standard state-space models is that the
observation error has no clear interpretation. The use of a mean zero
Gaussian distribution implies that the observed data $X_t$ will be
higher than $N_t$ as often as it is lower than $N_t$. It is hard
to identify a mechanism that would cause such errors. A more likely
form of observation error, and one that has been recognized for well
over a century, results from failing to detect individuals that are
present. Imperfect detection may be attributable to
characteristics of the species under study, such as its elusiveness,
or to the failings of the ecologist collecting the data in the field.
Although a vast number of methods have been devised for accounting for
this form of observation error, rarely have these methods been
integrated into state-space models \citep[but
see][]{buckland_etal:2004}.

@@

A third problem with these models is that they do not admit
spatial variation, which
%. Recent work has sought to remedy this deficiencyThis is not a fair criticism because this
%lies beyond the scope of traditional state-space models, but
%it does
restrict the utility of their since it is typically
impossible assess the impacts of factors such as habitat
fragmentation or climate change on population dynamics without
considering spatial variation in abundance. Indeed many populations are
regulated by spatial processes such as source-sink dynamics.

Although numerous extensions have been proposed to allow for spatial
variation in abundance, in this paper we focus on the model of
\citet{dail_madsen:2011} developed a model (henceforth the DM model)
that simultaneous resolves each of these problems with traditional state-space
models. Specifically, the model (1) recognizes the discrete nature of
count data and the underlying true abundance state, (2) has estimable
parameters, including those describing the initial state at time
$t=1$, and (3) allows for inference about both temporal and spatial
variation in abundance.
The DM model allows for inference about spatial and temporal
dynamics in abundance while accounting for observation error using
only spatially and temporally replicated count data.  Their model is
an extension of the closed-population $N$-mixture model
\citep{royle:2004biom}, which was designed specifically to address the
problem of
modeling spatial variation in abundance in the presence of observation
error, specifically, when detection probability is less than one. The
DM extension relaxes the assumption of population
closure and includes explicit parameters describing population change
as a first-order Markovian process. [more details]

[applications / potential etc... ] The DM model is relevant to a vast
number of problems confronting basic and applied ecological
research. Answering questions about climate change etc... require
spatially and temporally extensive datasets such as result from the
North American Breeding Bird Survey (BBS)
\citep{robbins_etal:1986}. However, most of the datasets of the
appropriate scale are
collected by volunteers and using protocols that are not amenable to
traditional approaches of modeling abundance and detection probability
based on standard capture-recapture methods. Instead they have, at
best, simple count data for which, when it can be interpreted as
counts of unique individuals, the DM model can be applied.. Yadi ya
